---
title: "Mendelian Randomization analysis of DeepSEA generated data"
output: html_notebook
---
Let's start by importing the Mendelian Randomization package we're going to use and the always helpful `dbplyr` and `tidyverse` packages.
```{r}
library(MendelianRandomization)
library(dplyr)
library(tidyverse)
```

First let's load our data into a dataframe. This dataframe has 5 non-null columns: 
1. `seq_num`
2. `X_pred_mean`: Effect size for the \( Z \rightarrow X \) relationship. Determined by taking the difference between predicted transcription factor binding probability for a reference sequence and a mutated version of the reference.
3. `X_pred_var`: Standard error of the \( Z \rightarrow X \) effect size estimate.
4. `Y_pred_mean`: Effect size for the \( Z \rightarrow Y \) relationship. Determined by taking the difference between predicted chromatin accessibility probability for a reference sequence and a mutated version of the reference.
5. `Y_pred_var`: Standard error of the \( Z \rightarrow Y \) effect size estimate.

```{r}
data_dir = "../../dat"
seq_predictions = read.csv(file.path(data_dir, "means_and_uncertainties_new_code_2.csv"))
seq_predictions
```

Before we do anything else, we have to make our data compatible with MR-Egger regression. MR-Egger requires that we change the signs of all of our \( X \rightarrow Y \) effect sizes to be positive. Let's do that now.
```{r}
seq_predictions %>%
  mutate(
    is.negative = X_pred_mean < 0,
    X_pred_mean = ifelse(is.negative, -X_pred_mean, X_pred_mean),
    Y_pred_mean = ifelse(is.negative, -Y_pred_mean, Y_pred_mean)
  )
```


```{r}
ivw_vals = list()
egger_vals = list()

egger_result <- matrix(ncol=4, nrow=25)
ivw_result <- matrix(ncol=3, nrow=25)


for (seq in (1:25)) {
  seq_predictions = subset(seq_predictions_all, seq_num == seq)

  bxt <- unlist(seq_predictions["X_pred_mean"])
  bxset <- unlist(seq_predictions["X_pred_var"])
  byt <- unlist(seq_predictions["Y_pred_mean"])
  byset <- unlist(seq_predictions["Y_pred_var"])
  MRInputObject <- mr_input(bx = bxt,
                            bxse = bxset,
                            by = byt,
                            byse = byset)
  EggerObject <- mr_egger(MRInputObject)
  IVWObject <- mr_ivw(MRInputObject)
  
  egger_result[seq, ] <- c(EggerObject$Estimate, EggerObject$CILower.Est, EggerObject$CIUpper.Est, EggerObject$Pleio.pval)
  ivw_result[seq, ] <- c(IVWObject$Estimate, IVWObject$CILower, IVWObject$CIUpper)
}

colnames(egger_result) <- c("ce", "cil", "ciu", "pleio")
egger_result <- as_tibble(egger_result)
colnames(ivw_result) <- c("ce", "cil", "ciu")
ivw_result <- as_tibble(ivw_result)
```


Now we have the causal effect estimates, confidence intervals, and (for Egger) pleiotropy p-values.
```{r}
ivw_result
egger_result
```

Let's summarize the IVW & Egger causal effect estimates as histograms. This will give us an idea of how heterogeneous the alleged causal relationships are at different regions of the genome.
```{r}
qplot(ce, data=ivw_result, binwidth=.02)
qplot(ce, data=egger_result, binwidth=.02)
```

Let's also look at the distribution of pleiotropy p-values for different sequences. If they're concentrated around .5, this means that the algorithm believes there to be very little pleiotropy.

```{r}
qplot(pleio, data=egger_result, binwidth=.04)
```
